\textbf{Microbiota predict \emph{Clostridium difficile} severity in germ-free mice colonized with human feces}

### 2022-01-23 submission/manuscript.tex
proofread coathur edits

The severity of \emph{Clostridioides difficile} infections (CDI) has
increased over the last few decades. Patient age, white blood cell
count, creatinine levels as well as \emph{C. difficile} ribotype and
toxin genes have been associated with disease severity. However, it is
unclear whether there is an association between the gut microbiota and
disease severity. The gut microbiota is known to interact with \emph{C.
difficile} during infection. Perturbations to the gut microbiota are
necessary for \emph{C. difficile} to colonize the gut. The gut
microbiota is thought to impair \emph{C. difficile} colonization through
bile acid metabolism, nutrient consumption and bacteriocin production.
Here we sought to demonstrate that members of the gut bacterial
communities can contribute to disease severity. We derived diverse gut
communities by colonizing germ-free mice with different human fecal
communities. The mice were then infected with a single \emph{C.
difficile} ribotype 027 clinical isolate which resulted in moribundity
and histopathologic differences. The variation in severity was
associated with the human fecal community that the mice received.
Generally, bacterial populations with pathogenic potential, such as
\emph{Escherichia}, \emph{Helicobacter}, and \emph{Klebsiella}, were
associated with more severe outcomes. Bacterial groups associated with
fiber degradation, bile acid metabolism and lantibiotic production, such
as \emph{Anaerostipes} and \emph{Coprobacillus}, were associated with
less severe outcomes. These data indicate that gut bacteria can
influence CDI disease severity.

@@

\emph{Clostridioides difficile} colonization can result in no symptoms
or develop into an infection, ranging in severity from mild diarrhea to
toxic megacolon, sepsis and death. Models that predict severity and
guide treatment decisions are based on clinical factors and \emph{C.
difficile} characteristics. Although the gut microbiome plays a role in
protecting against CDI, it has not been investigated for its effect on
CDI disease severity or incorporated into attempts to predict disease
severity. We demonstrated that variation in the microbiome of mice
colonized with human feces yielded a range of disease outcomes. These
results revealed groups of bacteria associated with both severe and mild
\emph{C. difficile} infection outcomes. Gut bacterial community data
from patients with CDI could improve our ability to identify patients at
risk of developing more severe disease and improve interventions which
target \emph{C. difficile} and the gut bacteria to reduce host damage.

@@

\emph{Clostridioides difficile} infections (CDI) have increased in
incidence and severity since \emph{C. difficile} was first identified as
the cause of antibiotic-associated pseudomembranous colitis (1). CDI
disease severity can range from asymptomatic to mild diarrhea to toxic
megacolon and death. The Infectious Diseases Society of America (IDSA)
and Society for Healthcare Epidemiology of America (SHEA) guidelines
defines severe CDI in terms of a white blood cell count greater than
15,000 cells/mm\(^3\) and/or a serum creatinine greater than 1.5 mg/dL.
Patients who develop shock or hypotension, ileus, or toxic megacolon are
considered to have fulminant CDI (2). Since these measures are outcomes
of CDI, these measures have limited ablity to predict risk of severe CDI
when CDI is fiest detected. Schemes have been developed to score a
patient's risk for severe CDI outcomes based on clinical factors but
have not been robust for broad application (3). Thus, we have limited
ability to prevent patients from developing severe CDI. Ideally, we
could identify patients that are high-risk of severe disease so their
treatments could be adjusted to alter the clinical trajectory.

@@

Here, we sought to determine whether the gut bacteria contribute to
disease severity. We hypothesized that since the gut bacteria affect the
metabolism of \emph{C. difficile} and its infection dynamics, it can
also affect the disease severity of the infection. To test this
hypothesis, we colonized germ-free C57BL/6 mice with human fecal samples
to create varied gut communities. We then challenged the mice with a
single \emph{C. difficile} isolate, ribotype 027 (RT027) clinical
isolate 431, and followed the mice for the development of severe
outcomes of moribundity and histopathologic cecal tissue damage. Since
the murine host and \emph{C. difficile} isolate were the same and only
the gut community varied, the variation in disease severity we observed
was attributable to the microbiome.

@@

\textbf{\emph{C. difficile} is able to infect germ-free mice colonized
with human fecal microbial communities without antibiotics} To produce
gut microbiomes with greater variation than those found in conventional
mouse colonies, we colonized germ-free mice with bacteria from human
feces (14). We inoculated germ-free C57BL/6 mice with homogenized feces
from one of 15 human fecal samples via oral gavage. These human fecal
samples were selected because they represented diverse community
structures based on community clustering (15). The gut communities were
allowed to equlibriate for two weeks post inoculation (16). We then
surveyed the bacterial members of the gut communities by 16S rRNA gene
sequencing of murine fecal pellets (Figure 1A). The bacterial
communities from each mouse grouped more closely to those communities
from mice that received the same human fecal donor community than to the
mice who received a different human fecal donor community (Figure 1B).
The communities were primarily composed of populations of
\emph{Clostridia}, \emph{Bacteroidia}, \emph{Erysipelotrichia},
\emph{Bacilli}, and \emph{Gammaproteobacteria}. However, the gut
bacterial communities of each donor group of mice harbored unique
relative abundance distributions of the shared bacterial classes.

Next, we tested this set of mice with their human-derived gut microbial
communities for susceptibility to \emph{C. difficile} infection. A
typical mouse model of CDI requires pre-treatment of conventional mice
with antibiotics, such as clindamycin, to become susceptible to \emph{C.
difficile} colonization (17, 18). However, we wanted to avoid modifying
the gut communities with an antibiotic to maintain their unique
microbial compositions and ecological relationships. Since some of these
communities came from people at increased risk of CDI, for example due
to hospitalization or recent antibiotic use (15), we decided to test if
\emph{C. difficile} was able to infect these mice without any antibiotic
perturbation. We hypothesized that \emph{C. difficile} would be able to
colonize the mice who received their gut communities from a donor with
perturbed community. Mice were challenged with 10\(^{3}\) \emph{C.
difficile} RT027 clinical isolate 431 spores. The mice were followed for
10 days post-challenge, and their stool was collected and plated for
\emph{C. difficile} colony forming units (CFU) to determine the extent
of the infection. Surprisingly, communities from all donors were able to
be colonized (Figure 2). Two mice were able to resist \emph{C.
difficile} colonization, both received their community donor N3, which
may be attributed to experimental variation since this group also had
more mice. By colonizing germ-free mice with different human fecal
communities, we were able to generate diverse gut communities in mice,
which were susceptible to \emph{C. difficile} infection without further
modification of the gut community.

\textbf{Infection severity varies by initial community.} After we
challenged the mice with \emph{C. difficile}, we investigated the
outcome from the infection and its relationship to the initial
community. We followed the mice for 10 days post-challenge for
colonization density, toxin production, and mortality. Seven mice, from
Donors N1, N2 and N5, were not colonized at detectable levels on the day
after \emph{C. difficile} challenge but were infected
(\textgreater10\(^{6}\)) by the end of the experiment. All mice that
received their community from Donor M1 through M6 succumbed to the
infection and became moribund within 3 days post-challenge. The
remaining mice, except the uninfected Donor N3 mice, maintained \emph{C.
difficile} infection through the end of the experiment (Figure 2). At 10
days post-challenge, or earlier for the moribund mice, mice were
euthanised and fecal material were assayed for toxin activity and cecal
tissue was collected and scored for histopathologic signs of disease
(Figure 3). Overall, there was greater toxin activity detected in the
stool of the moribund mice (\emph{P} = 0.003). However, when looking at
each group of mice, we observed a range in toxin activity for both the
moribund and non-moribund mice (Figure 3A). Non-moribund mice from
Donors N4 through N9 had comparable toxin activity as the moribund mice.
Additionally, not all moribund mice had toxin activity detected in their
stool. Next we examined the cecal tissue for histopathologic damage.
Moribund mice had high levels of epithelial damage, tissue edema, and
inflammation (Figure S1) similar to previously reported histopathologic
findings for \emph{C. difficile} RT027 (19). As observed with toxin
activity, the moribund mice had higher histopathologic scores than the
non-moribund mice (\emph{P} \textless{} 0.001). However, unlike the
toxin activity, all moribund mice had consistently high histopathologic
summary scores (Figure 3B). The non-moribund mice, Donor groups N1
through N9, had a range in tissue damage from none detected to similar
levels as the moribund mice, which grouped by community donor. Together,
the toxin activity, histopathologic score, and moribundity showed
variation across the donor groups but were largely consistent within
each donor group.

\textbf{Microbial community members explain variation in CDI severity.}
We next interrogated the bacterial communities at the time of \emph{C.
difficile} challenge (day 0) for their relationship to infection
outcomes using linear discriminant analysis (LDA) effect size (LEfSe)
analysis to identify individual bacterial populations that could explain
the variation in disease severity. We split the mice into groups by
severity level based on their moribundity and histopathologic score. We
dichotomized the histopathologic scores into high and low groups by
splitting on the median score of 5. This analysis revealed 20 genera
that were significantly different by the disease severity (Figure 4A).
Bacterial genera \emph{Turicibacter}, \emph{Streptococcus},
\emph{Staphylococcus}, \emph{Pseudomonas}, \emph{Phocaeicola},
\emph{Parabacteroides}, \emph{Bacteroides}, and
\emph{Escherichia/Shigella} were detected at higher relative abundances
in the mice that became moribund. Populations of \emph{Anaerotignum},
\emph{Coprobacillus}, \emph{Enterocloster}, and \emph{Murimonas} were
more abundant in the non-moribund mice that would develop only low
intestinal injury. To understand the role of toxin activity in disease
severity, we applied LEfSe to identify the genera most likely to explain
the differences between the presence and absence of detected toxin
activity (Figure 4B). Many genera that associated with toxin production
were also associated with moribundity, such as populations of
\emph{Escherichia/Shigella} and \emph{Bacteroides}. Likewise, there were
genera such as \emph{Anaerotignum}, \emph{Enterocloster}, and
\emph{Murimonas} that were associated with no detected toxin production
that also exhibited greater relative abundance in communities from
non-moribund mice with a low histopathologic score. Lastly, we tested
for correlations between the endpoint relative abundances of bacterial
operational taxonomic units (OTUs) and the histopathologic summary score
(Figure 4C). The endpoint relative abundance of \emph{Bacteroides} was
positively correlated with histopathologic score, as its day 0 relative
abundance did with disease severity (Figure 4A). Populations of
\emph{Klebsiella} and \emph{Prevotellaceae} were positively correlated
with the histopathologic score and were increased in the group of mice
with detectable toxin. This analysis identified bacterial genera that
were associated with the variation in moribundity, histopathologic
score, and toxin activity of CDI.

We next determined whether collectively, bacterial community membership
and relative abundance could be predictive of the CDI disease outcome.
We trained random forest models with bacterial community relative
abundance data from the day of colonization at each taxonomic rank to
predict toxin production, moribundity, and day 10 post-challenge
histopathologic summary score. For predicting if toxin would be
produced, microbial populations aggregated by phylum rank classification
performed similarly as models using lower taxonomic ranks (AUROC = 0.83,
Figure S2). \emph{C. difficile} was more likely to produce toxin when
the community it infected had less abundant populations of
\emph{Verrucomicrobia} and \emph{Campilobacterota} and had more abundant
populations of \emph{Proteobacteria} (Figure 5A). Next, we assessed the
ability of the community to predict moribundity. Bacteria grouped by
class rank classification was sufficient to predict which mice would
succumb to the infection before the end of the experiment (AUROC = 0.91,
Figure S2). The features with the greatest effect showed that
communities with greater populations of bacteria belonging to
\emph{Bacilli} and \emph{Firmicutes} and reduced populations of
\emph{Erysipelotrichia} were more likely to result in moribundity
(Figure 5B). Only one other class of bacteria was decreased in moribund
mice, a group of unclassified \emph{Clostridia}. Lastly, the relative
abundances of genera were able to predict a high or low histopathologic
score (histopathologic scores were dichotomized as in previous analysis,
AUROC = 0.99, Figure S2). No genera had a significantly greater effect
on the model performance than any others, indicating the model was
reliant on many genera for the correct prediction. The model used some
of the genera identified in the LEfSe analysis, such as
\emph{Coprobacillus}, \emph{Anaerostipes}, and \emph{Hungatella}.
Communities with greater abundances of \emph{Hungatella},
\emph{Eggerthella}, \emph{Bifidobacterium}, \emph{Duncaniella} and
\emph{Neisseria} were more likely to have high histopathologic scores.
These models have shown that the relative abundance of bacterial
populations and their relationship to each other could be used to
predict the variation in moribundity, histopathologic score, and toxin
activity of CDI.

@@

Challenging mice colonized with different human fecal communities with a
single \emph{C. difficile} isolate, RT027 isolate 431, demonstrated that
variation in the gut bacteria affects the \emph{C. difficile} infection
disease severity. Our analysis revealed an association between the
relative abundance of bacterial community members and disease severity.
Previous studies investigating the severity of CDI disease involving the
microbiome have had limited ability to interrogate this relationship
between the microbiome and disease severity. Studies that have used
clinical data have limited ability to control variation in the host,
microbiome or \emph{C. difficile} ribotype (20). Murine experiments
typically use a single mouse colony and different \emph{C. difficile}
ribotypes to create severity differences (21). Recently, our group has
begun uncovering the effect microbiome variation has on \emph{C.
difficile} infection. We showed the variation in the bacterial
communities between mice from different mouse colonies resulted in
different clearance rates of \emph{C. difficile} (22). We also showed
varied ability of mice to spontaneously eliminate \emph{C. difficile}
infection when they were treated with different antibiotics prior to
\emph{C. difficile} challenge (23). Overall, the results presented here
have demonstrated that the gut bacterial community contributed to the
severity of \emph{C. difficile} infection.

\emph{C. difficile} can lead to asymptomatic colonization or infections
with severity ranging from mild diarrhea to death. Physicians use
classification tools to identify patients most at risk of developing a
severe infection using white blood cell counts, serum albumin level, or
serum creatinine level (2, 24, 25). Those levels are driven by the
activities in the intestine (26). Research into the drivers of this
variation have revealed factors that make \emph{C. difficile} more
virulent. Strains are categorized for their virulence by the presence
and production of the toxins TcdA, TcdB, and binary toxin and the
prevalence in outbreaks, such as ribotypes 027 and 078 (17, 27--30).
However, other studies have shown that disease is not necessarily linked
with toxin production (31) or the strain (32). Furthermore, there is
variation in the genome, growth rate, sporulation, germination, and
toxin production in different isolates of a strain (33--36). This
variation may help explain why severe CDI prediction tools often miss
identifying many patients with CDI that will develop severe disease (3,
21, 37, 38). Therefore, it is necessary to gain a full understanding of
all factors contributing to disease variation to improve our ability to
predict severity.

The state of the gut bacterial community determines the ability of
\emph{C. difficile} to colonize and persist in the intestine. \emph{C.
difficile} is unable to colonize an unperturbed healthy gut community
and is only able to become established after a perturbation (18). Once
colonized, the different communities lead to different metabolic
responses and dynamics of the \emph{C. difficile} population (8, 23,
39). Gut bacteria metabolize primary bile acids into secondary bile
acids (40, 41). The concentration of these bile acids affects
germination, growth, toxin production and biofilm formation (9, 10, 42,
43). Members of the bacterial community also affect other metabolites
\emph{C. difficile} utilizes. \emph{Bacteroides thetaiotaomicron}
produce sialidases which release sialic acid from the mucosa for
\emph{C. difficile} to utilize (44, 45). The nutrient environment
affects toxin production (46). Thus, many of the actions of the gut
bacteria modulate \emph{C. difficile} in ways that could affect the
infection and resultant disease.

>>>>

The severity of \emph{Clostridioides difficile} infections (CDI) has
increased over the last few decades. Patient age, white blood cell
count, creatinine levels as well as \emph{C. difficile} ribotype and
toxin genes have been associated with disease severity. However, it is
unclear whether there is an association between members of the gut
microbiota and disease severity. The gut microbiota is known to interact
with \emph{C. difficile} during infection. Perturbations to the gut
microbiota are necessary for \emph{C. difficile} to colonize the gut.
The gut microbiota can inhibit \emph{C. difficile} colonization through
bile acid metabolism, nutrient consumption and bacteriocin production.
Here we sought to demonstrate that members of the gut bacterial
communities can also contribute to disease severity. We derived diverse
gut communities by colonizing germ-free mice with different human fecal
communities. The mice were then infected with a single \emph{C.
difficile} ribotype 027 clinical isolate which resulted in moribundity
and histopathologic differences. The variation in severity was
associated with the human fecal community that the mice received.
Generally, bacterial populations with pathogenic potential, such as
\emph{Escherichia}, \emph{Helicobacter}, and \emph{Klebsiella}, were
associated with more severe outcomes. Bacterial groups associated with
fiber degradation, bile acid metabolism and lantibiotic production, such
as \emph{Anaerostipes} and \emph{Coprobacillus}, were associated with
less severe outcomes. These data indicate that, in addition to the host
and \emph{C. difficile}, populations of gut bacteria can influence CDI
disease severity.

@@

\emph{Clostridioides difficile} colonization can result in no symptoms
or develop into an infection, ranging in severity from mild diarrhea to
toxic megacolon, sepsis, and death. Models that predict severity and
guide treatment decisions are based on clinical factors and \emph{C.
difficile} characteristics. Although the gut microbiome plays a role in
protecting against CDI, it has its effect on CDI disease severity is
unclear and has not been incorporated into disease severity models. We
demonstrated that variation in the microbiome of mice colonized with
human feces yielded a range of disease outcomes. These results revealed
groups of bacteria associated with both severe and mild \emph{C.
difficile} infection outcomes. Gut bacterial community data from
patients with CDI could improve our ability to identify patients at risk
of developing more severe disease and improve interventions which target
\emph{C. difficile} and the gut bacteria to reduce host damage.

@@

\emph{Clostridioides difficile} infections (CDI) have increased in
incidence and severity since \emph{C. difficile} was first identified as
the cause of antibiotic-associated pseudomembranous colitis (1). CDI
disease severity can range from mild diarrhea to toxic megacolon and
death. The Infectious Diseases Society of America (IDSA) and Society for
Healthcare Epidemiology of America (SHEA) guidelines define severe CDI
in terms of a white blood cell count greater than 15,000 cells/mm\(^3\)
and/or a serum creatinine greater than 1.5 mg/dL. Patients who develop
shock or hypotension, ileus, or toxic megacolon are considered to have
fulminant CDI (2). Since these measures are CDI outcomes, they have
limited ability to predict risk of severe CDI when the infection is
first detected. Schemes have been developed to score a patient's risk
for severe CDI outcomes based on clinical factors but have not been
robust for broad application (3). Thus, we have limited ability to
prevent patients from developing severe CDI.

@@

Recently, Nagao-Kitamoto \emph{et al.} showed germ-free mice which
received a microbial community from different human donors resulted in
variation in a subsequent \emph{C. difficile} infection (14). Here, we
sought to further elucidate the relationship between members of the gut
bacterial community and CDI disease severity. We hypothesized that since
specific groups of gut bacteria affect the metabolism of \emph{C.
difficile} and its infection dynamics, we can also identify groups of
bacteria that affect the disease severity of the infection. To test this
hypothesis, we colonized germ-free C57BL/6 mice with human fecal samples
to create varied gut communities. We then challenged the mice with
\emph{C. difficile} ribotype 027 (RT027) and followed the mice for the
development of severe outcomes of moribundity and histopathologic cecal
tissue damage. Since the murine host and \emph{C. difficile} isolate
were the same and only the gut community varied, the variation in
disease severity we observed was attributable to the gut microbiome.

@@

\textbf{\emph{C. difficile} is able to infect germ-free mice colonized
with human fecal microbial communities without antibiotics} To produce
gut microbiomes with greater variation than those found in conventional
mouse colonies, we colonized germ-free mice with bacteria from human
feces (15). We inoculated germ-free C57BL/6 mice with homogenized feces
from one of 15 human fecal samples via oral gavage. These human fecal
samples were selected because they represented diverse community
structures based on community clustering (16). The gut communities were
allowed to equilibrate for two weeks post-inoculation (17). We then
surveyed the bacterial members of the gut communities by 16S rRNA gene
sequencing of murine fecal pellets (Figure 1A). The bacterial
communities from each mouse grouped more closely to those communities
from mice that received the same human fecal donor community than to the
mice who received a different human fecal donor community (Figure 1B).
The communities were primarily composed of populations of
\emph{Clostridia}, \emph{Bacteroidia}, \emph{Erysipelotrichia},
\emph{Bacilli}, and \emph{Gammaproteobacteria}. However, the gut
bacterial communities of each donor group of mice harbored unique
relative abundance distributions of the shared bacterial classes.

Next, we tested this set of mice with their human-derived gut microbial
communities for susceptibility to \emph{C. difficile} infection. A
typical mouse model of CDI requires pre-treatment of conventional mice
with antibiotics, such as clindamycin, to become susceptible to \emph{C.
difficile} colonization (18, 19). However, we wanted to avoid modifying
the gut communities with an antibiotic to maintain their unique
microbial compositions and ecological relationships. Since some of these
communities came from people at increased risk of CDI, such as recent
hospitalization or antibiotic use (16), we decided to test if \emph{C.
difficile} was able to infect these mice without any antibiotic
perturbation. We hypothesized that \emph{C. difficile} would be able to
colonize the mice who received their gut communities from a donor with a
perturbed community. Mice were challenged with 10\(^{3}\) \emph{C.
difficile} RT027 clinical isolate spores. The mice were followed for 10
days post-challenge, and their stool was collected and plated for
\emph{C. difficile} colony forming units (CFU) to determine the extent
of the infection. Surprisingly, communities from all donors were able to
be colonized (Figure 2). Two mice were able to resist \emph{C.
difficile} colonization, both received their community donor N1, which
may be attributed to experimental variation since this group also had
more mice. By colonizing germ-free mice with different human fecal
communities, we were able to generate diverse gut communities in mice,
which were susceptible to \emph{C. difficile} infection without further
modification of the gut community.

\textbf{Infection severity varies by initial community.} After we
challenged the mice with \emph{C. difficile}, we investigated the
outcome from the infection and its relationship to the initial
community. We followed the mice for 10 days post-challenge for
colonization density, toxin production, and mortality. Seven mice, from
Donors N1, N3, N4, and N5, were not colonized at detectable levels on
the day after \emph{C. difficile} challenge but were infected
(\textgreater10\(^{6}\)) by the end of the experiment. All mice that
received their community from Donor M1 through M6 succumbed to the
infection and became moribund within 3 days post-challenge. The
remaining mice, except the uninfected Donor N1 mice, maintained \emph{C.
difficile} infection through the end of the experiment (Figure 2). At 10
days post-challenge, or earlier for the moribund mice, mice were
euthanised and fecal material were assayed for toxin activity and cecal
tissue was collected and scored for histopathologic signs of disease
(Figure 3). Overall, there was greater toxin activity detected in the
stool of the moribund mice (\emph{P} = 0.003). However, when looking at
each group of mice, we observed a range in toxin activity for both the
moribund and non-moribund mice (Figure 3A). Non-moribund mice from
Donors N2 and N5 through N9 had comparable toxin activity as the
moribund mice. Additionally, not all moribund mice had toxin activity
detected in their stool. Next, we examined the cecal tissue for
histopathologic damage. Moribund mice had high levels of epithelial
damage, tissue edema, and inflammation (Figure S1) similar to previously
reported histopathologic findings for \emph{C. difficile} RT027 (20). As
observed with toxin activity, the moribund mice had higher
histopathologic scores than the non-moribund mice (\emph{P} \textless{}
0.001). However, unlike the toxin activity, all moribund mice had
consistently high histopathologic summary scores (Figure 3B). The
non-moribund mice, Donor groups N1 through N9, had a range in tissue
damage from none detected to similar levels as the moribund mice, which
grouped by community donor. Together, the toxin activity,
histopathologic score, and moribundity showed variation across the donor
groups but were largely consistent within each donor group.

\textbf{Microbial community members explain variation in CDI severity.}
We next interrogated the bacterial communities at the time of \emph{C.
difficile} challenge (day 0) for their relationship to infection
outcomes using linear discriminant analysis (LDA) effect size (LEfSe)
analysis to identify individual bacterial populations that could explain
the variation in disease severity. We split the mice into groups by
severity level based on their moribundity and histopathologic score. We
dichotomized the histopathologic scores into high and low groups by
splitting on the median score of 5. This analysis revealed 20 genera
that were significantly different by the disease severity (Figure 4A).
Bacterial genera \emph{Turicibacter}, \emph{Streptococcus},
\emph{Staphylococcus}, \emph{Pseudomonas}, \emph{Phocaeicola},
\emph{Parabacteroides}, \emph{Bacteroides}, and
\emph{Escherichia/Shigella} were detected at higher relative abundances
in the mice that became moribund. Populations of \emph{Anaerotignum},
\emph{Coprobacillus}, \emph{Enterocloster}, and \emph{Murimonas} were
more abundant in the non-moribund mice that would develop only low
intestinal injury. To understand the role of toxin activity in disease
severity, we applied LEfSe to identify the genera most likely to explain
the differences between the presence and absence of detected toxin
activity (Figure 4B). Many genera that associated with the presence of
toxin were also associated with moribundity, such as populations of
\emph{Escherichia/Shigella} and \emph{Bacteroides}. Likewise, there were
genera such as \emph{Anaerotignum}, \emph{Enterocloster}, and
\emph{Murimonas} that were associated with no detected toxin that also
exhibited greater relative abundance in communities from non-moribund
mice with a low histopathologic score. Lastly, we tested for
correlations between the endpoint relative abundances of bacterial
operational taxonomic units (OTUs) and the histopathologic summary score
(Figure 4C). The endpoint relative abundance of \emph{Bacteroides} was
positively correlated with histopathologic score, as its day 0 relative
abundance did with disease severity (Figure 4A). Populations of
\emph{Klebsiella} and \emph{Prevotellaceae} were positively correlated
with the histopathologic score and were increased in the group of mice
with detectable toxin. This analysis identified bacterial genera that
were associated with the variation in moribundity, histopathologic
score, and toxin.

We next determined whether, collectively, bacterial community membership
and relative abundance could be predictive of the CDI disease outcome.
We trained random forest models with bacterial community relative
abundance data from the day of colonization at each taxonomic rank to
predict toxin, moribundity, and day 10 post-challenge histopathologic
summary score. For predicting if detectable toxin would be produced,
microbial populations aggregated by phylum rank classification performed
similarly as models using lower taxonomic ranks (AUROC = 0.83, Figure
S2). \emph{C. difficile} was more likely to produce detectable toxin
when the community infected had less abundant populations of
\emph{Verrucomicrobia} and \emph{Campilobacterota} and had more abundant
populations of \emph{Proteobacteria} (Figure 5A). Next, we assessed the
ability of the community to predict moribundity. Bacteria grouped by
class rank classification was sufficient to predict which mice would
succumb to the infection before the end of the experiment (AUROC = 0.91,
Figure S2). The features with the greatest effect showed that
communities with greater populations of bacteria belonging to
\emph{Bacilli} and \emph{Firmicutes} and reduced populations of
\emph{Erysipelotrichia} were more likely to result in moribundity
(Figure 5B). Only one other class of bacteria was decreased in moribund
mice, a group of unclassified \emph{Clostridia}. Lastly, the relative
abundances of genera were able to predict a high or low histopathologic
score (histopathologic scores were dichotomized as in previous analysis,
AUROC = 0.99, Figure S2). No genera had a significantly greater effect
on the model performance than any others, indicating the model was
reliant on many genera for the correct prediction. The model used some
of the genera identified in the LEfSe analysis, such as
\emph{Coprobacillus}, \emph{Anaerostipes}, and \emph{Hungatella}.
Communities with greater abundances of \emph{Hungatella},
\emph{Eggerthella}, \emph{Bifidobacterium}, \emph{Duncaniella} and
\emph{Neisseria} were more likely to have high histopathologic scores.
These models have shown that the relative abundance of bacterial
populations and their relationship to each other could be used to
predict the variation in moribundity, histopathologic score, and
detectable toxin of CDI.

@@

Challenging mice colonized with different human fecal communities with
\emph{C. difficile} RT027 demonstrated that variation in members of the
gut microbiome affects \emph{C. difficile} infection disease severity.
Our analysis revealed an association between the relative abundance of
bacterial community members and disease severity. Previous studies
investigating the severity of CDI disease involving the microbiome have
had limited ability to interrogate this relationship between the
microbiome and disease severity. Studies that have used clinical data
have limited ability to control variation in the host, microbiome or
\emph{C. difficile} ribotype (21). Murine experiments typically use a
single mouse colony and different \emph{C. difficile} ribotypes to
create severity differences (22). Recently, our group has begun
uncovering the effect microbiome variation has on \emph{C. difficile}
infection. We showed the variation in the bacterial communities between
mice from different mouse colonies resulted in different clearance rates
of \emph{C. difficile} (23). We also showed varied ability of mice to
spontaneously eliminate \emph{C. difficile} infection when they were
treated with different antibiotics prior to \emph{C. difficile}
challenge (24). Overall, the results presented here have demonstrated
that the gut bacterial community contributed to the severity of \emph{C.
difficile} infection.

\emph{C. difficile} can lead to asymptomatic colonization or infections
with severity ranging from mild diarrhea to death. Physicians use
classification tools to identify patients most at risk of developing a
severe infection using white blood cell counts, serum albumin level, or
serum creatinine level (2, 25, 26). Those levels are driven by the
activities in the intestine (27). Research into the drivers of this
variation have revealed factors that make \emph{C. difficile} more
virulent. Strains are categorized for their virulence by the presence
and production of the toxins TcdA, TcdB, and binary toxin and the
prevalence in outbreaks, such as ribotypes 027 and 078 (18, 28--31).
However, other studies have shown that disease is not necessarily linked
with toxin production (32) or the strain (33). Furthermore, there is
variation in the genome, growth rate, sporulation, germination, and
toxin production in different isolates of a strain (34--37). This
variation may help explain why severe CDI prediction tools often miss
identifying many patients with CDI that will develop severe disease (3,
22, 38, 39). Therefore, it is necessary to gain a full understanding of
all factors contributing to disease variation to improve our ability to
predict severity.

The state of the gut bacterial community determines the ability of
\emph{C. difficile} to colonize and persist in the intestine. \emph{C.
difficile} is unable to colonize an unperturbed healthy gut community
and is only able to become established after a perturbation (19). Once
colonized, the different communities lead to different metabolic
responses and dynamics of the \emph{C. difficile} population (8, 24,
40). Gut bacteria metabolize primary bile acids into secondary bile
acids (41, 42). The concentration of these bile acids affects
germination, growth, toxin production and biofilm formation (9, 10, 43,
44). Members of the bacterial community also affect other metabolites
\emph{C. difficile} utilizes. \emph{Bacteroides thetaiotaomicron}
produce sialidases which release sialic acid from the mucosa for
\emph{C. difficile} to utilize (45, 46). The nutrient environment
affects toxin production (47). Thus, many of the actions of the gut
bacteria modulate \emph{C. difficile} in ways that could affect the
infection and resultant disease.

A myriad of studies have explored the relationship between the
microbiome and CDI disease. CDI studies often use an infection model
with a single homogeneous bacterial community. Collins \emph{et al.}
used multiple human communities to colonize mice, however the
communities were pooled prior to gavaging into germ-free mice (48),
resulting in a single human-derived community. Studies examining
difference in disease often use different \emph{C. difficile} strains or
ribotypes in mice with similar microbiota as a proxy for variation in
disease, such as strain 630 for non-severe and RT027 for severe (18, 28,
29, 49). Studies have also demonstrated variation in infection through
tapering antibiotic dosage (19, 24, 50) or by reducing the amount of
\emph{C. difficile} cells or spores used for the challenge (18, 50).
These studies either lack variation in the initial microbiome or have
variation in the \emph{C. difficile} infection itself, confounding any
association between variation in severity and the microbiome.

Recently, Nagao-Kitamoto \emph{et al.} demonstrated when only allowing
the microbiome to vary in a murine \emph{C. difficile} infection
experiemnt that the resultant disease variation associated with the
microbiome communities (14). Our data have demonstrated gut bacterial
relative abundances associate with variation in toxin production,
histopathologic scoring of the cecal tissue and mortality. This analysis
revealed populations of \emph{Akkermansia}, \emph{Anaerostipes},
\emph{Coprobacillus}, \emph{Enterocloster}, \emph{Lactonifactor}, and
\emph{Monoglobus} were more abundant in the microbiome of non-moribund
mice which had low histopathologic scores and no detected toxin. The
protective role of these genera are supported by previous studies.
\emph{Coprobacillus}, \emph{Lactonifactor}, and \emph{Monoglobus} have
been shown to be involved in dietary fiber fermentation and associated
with healthy communities (51--54). \emph{Anaerostipes} and
\emph{Coprobacillus}, which produce short chain fatty acids, have been
associated with healthy communities (55--57). Furthermore,
\emph{Coprobacillus}, which was abundant in mice with low
histopathologic scores but rare in all other mice, has been shown to
contain a putative type I lantibiotic gene cluster and inhibit \emph{C.
difficile} colonization (58--60). \emph{Akkermansia} and
\emph{Enterocloster} were also identified as more abundant in mice which
had a low histopathologic scores but have contradictory supporting
evidence in the current literature. In our data, \emph{Akkermansia} was
most abundant in the non-moribund mice with low histopathologic scores
but there were some moribund mice which had increased populations of
\emph{Akkermansia}. This could be attributed to either a more protective
mucus layer was present inhibiting colonization (60, 61) or mucus
consumption by \emph{Akkermansia} could have been crossfeeding \emph{C.
difficile} or exposing a niche for \emph{C. difficile} (62--64).
Similarly, \emph{Enterocloster} was more abundant and associated with
low histopathologic scores. It has been associated with healthy
populations and has been used to mono-colonize germ-free mice to reduce
the ability of \emph{C. difficile} to colonize (65, 66). However,
\emph{Enterocloster} has also been involved in infections, such as
bacteremia (67, 68). These data have exemplified populations of bacteria
that have the potential to be either protective or harmful. Thus, the
disease outcome is not likely based on the abundance of individual
populations of bacteria, rather it is the result of the interactions of
the community.

The groups of bacteria that were associated with either a higher
histopathologic score or moribundity are members of the indigenous gut
community that also have been associated with disease, often referred to
as opportunistic pathogens. Many of the populations with pathogenic
potential that associated with worse outcomes are also facultative
anaerobes. \emph{Enterococcus}, \emph{Klebsiella},
\emph{Shigella/Escherichia}, \emph{Staphylococcus}, and
\emph{Streptococcus} have been shown to expand after antibiotic use (16,
69, 70) and are commonly detected in CDI cases (71--74). In addition to
these populations, \emph{Eggerthella}, \emph{Prevotellaceae} and
\emph{Helicobacter}, which associated with worse outcomes, have also
been associated with intestinal inflammation (75--77). Recently,
\emph{Helicobacter hepaticus} was shown to be sufficient to cause
susceptibility to CDI in IL-10 deficient C57BL/6 mice (78). In our
experiments, when \emph{Helicobacter} was present, the infection
resulted in a high histopathologic score (Figure 4C). While we did not
use IL-10 deficient mice, it is possible the bacterial community or host
response are similarly modified by \emph{Helicobacter}, allowing
\emph{C. difficile} infection and host damage. Aside from
\emph{Helicobacter}, these groups of bacteria that associated with more
severe outcomes did not have a conserved association between their
relative abundance and the disease severity across all mice.
